---
title: "Healthcare"
output: pdf_document
---
```{r, include = F}
usr1 <- 'cuny'
pw <- 'msda'
```

```{r}
library(RMySQL)
mydb = dbConnect(MySQL(), user=usr1, password = pw, dbname = 'medicare')

dbSendQuery(mydb, 'USE medicare;')
query1 <- dbSendQuery(mydb, "
SELECT   m.nppes_provider_state, COUNT(*)
FROM     medicare.twentytwelve AS m
WHERE m.nppes_provider_country = 'US'
GROUP BY m.nppes_provider_state;")

temp <- fetch(query1)
temp
```

```{r}
names(temp) <- c('state', 'count')
library(ggplot2)
ggplot(temp, aes(x = state, y = count)) + geom_bar(stat = "identity", width = .75) + scale_y_log10() + coord_flip()
```

```{r}
query2 <- dbSendQuery(mydb, "
SELECT   m.hcpcs_code, COUNT(*)
FROM     medicare.twentytwelve AS m
WHERE m.nppes_provider_country = 'US'
GROUP BY m.hcpcs_code;")
temp2 <- fetch(query2)
temp2
names(temp2) <- c('code', 'count')

for(i in 1:3){
  temp3 <- temp2[temp2$count > 10000,]
  ggplot(temp3, aes(x = code, y = count)) + geom_bar(stat = "identity", width = .75) + coord_flip()
}
```

hcpcs are actually cpt codes, and they vary regionally (although most are constant)

```{r}
query3 <- dbSendQuery(mydb, "
+ SELECT   m.provider_type, COUNT(*)
+ FROM     medicare.twentytwelve AS m
+ WHERE m.nppes_provider_state = 'NY'
+ GROUP BY m.provider_type
+ ORDER BY COUNT;")
```

```{r}
query4 <- dbSendQuery(mydb, "
SELECT   m.hcpcs_code, COUNT(*)
FROM     medicare.twentytwelve AS m
WHERE m.nppes_provider_state = 'NY' AND m.provider_type = 'Anesthesiology'
GROUP BY m.hcpcs_code;")
temp5 <- fetch(query3)
temp5
names(temp5) <- c('type', 'count')

sort(temp4[,2])

ggplot(temp4, aes(x = type, y = count)) + geom_bar(stat = "identity", width = .75) + coord_flip()
```

```{r}
query5 <- dbSendQuery(mydb, "
SELECT   m.hcpcs_code, COUNT(DISTINCT m.npi)
FROM     medicare.twentytwelve AS m
WHERE m.nppes_provider_state = 'NY' AND m.provider_type = 'Anesthesiology'
GROUP BY m.hcpcs_code;")
temp <- fetch(query5)
names(temp) <- c('code', 'count')
temp[order(temp$count),]
```

67  01210    36
169 64495    39
93  01924    41
321 J3490    41
33  00630    43
46  00830    44
115 27096    46
59  00912    47
60  00914    47
139 62310    47
18  00520    51
1   00103    52
110 20610    52
288 99232    53
289 99233    53
225 93320    54
280 99212    54
22  00532    55
163 64484    56
168 64494    57
291 99291    58
167 64493    64
98  01936    65
12  00300    66
162 64483    67
21  00530    68
13  00320    69
70  01230    69
227 93325    70
83  01630    72
276 99203    72
31  00567    79
24  00537    81
157 64448    81
3   00140    84
35  00670    85
122 31500    85
150 64415    88
89  01844    91
193 76937    95
6   00145    99
282 99214   104
277 99204   106
196 77003   112
156 64447   121
80  01480   122
222 93312   122
76  01400   127
47  00840   132
101 01996   136
281 99213   136
2   00104   141
28  00562   141
140 62311   141
68  01214   144
92  01922   158
86  01810   165
287 99231   177
130 36556   184
229 93503   187
58  00910   231
100 01992   243
77  01402   290
194 76942   314
15  00400   396
40  00790   552
4   00142   753
132 36620   868
44  00810  1586
37  00740  1733

https://ocm.ama-assn.org/OCM/CPTRelativeValueSearchResults.do?locality=63&keyword=00740

00740 	Anesthesia for upper gastrointestinal endoscopic procedures, endoscope introduced proximal to duodenum
Nov 99.Volume 9, Issue 11, November 1999 

 	Arterial catheterization or cannulation for sampling, monitoring or transfusion (separate procedure); percutaneous
 	
44  00810  1400
37  00740  1569

```{r}
subset <- dbSendQuery(mydb, "
SELECT   *
FROM     medicare.twentytwelve AS m
WHERE m.nppes_provider_state = 'NY' AND m.provider_type = 'Anesthesiology';")
temp.sub <- fetch(subset)
